Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-07-23 demystify.tex
updates
services

          \item Increasing admission chance by being unique and standing out (\autoref{sec:improve-your-chance}).
          \item It takes about 10 minutes to review an application (\autoref{sec:ievaluate}).
          \item Sometimes adcom members dicuss applicants, but in most cases they make independent decisions (\autoref{sec:adcom-discuss}).           
          \item Adcom members only review applications assigned to them (typically matching their expertise) and rarely get involved in other applications (\autoref{sec:read-all}).

@@

In your SOP, focus on research motivation and convince us that you can achieve it through your experience, e.g., published papers. You should also back up your claims with concrete evidence. For example, if you say you have experience with teaching, then show what you did, e.g., undergrad TA or mentoring someone.  If you say you work on project X, then show some results (paper submitted (or even rejected), achieved X\% performance improvement over ..). Also, do not use more than two pages (references can be on the third page). Many SOP samples are \href{https://cs-sop.org/}{available here}.

@@

\begin{commentbox}[Vu:]
  I often read SOP and LORs first. If I am
  persuaded by then, I would skim over other factors and advocate for
  admission (unless I see red flags in other parts). However, if I am not
  convinced, then I will likely recommend rejection (unless I see
  something really stands out in other parts).
  \tcblower
  Do careful research on professors, don't mention \emph{emeritus} or adjunct faculty (see more about various types of faculty in \autoref{sec:faculty-types}).
  Also, be careful not to send statements to the wrong schools or mix
  facts (e.g., talking about school X but mentioning working with
  profs. at school Y; and do not talk about George Washington when applying to George Mason). I have seen such statements more times than I should.
\end{commentbox}

@@

\begin{commentbox}[Thanh:]
  Vietnamese universities typically offer specialized programs, such as the talented engineer program at HUST, that have highly competitive entrance exams and a limited number of available slots (e.g., 30 per year). However, these programs often set higher requirements for students, including more demanding tests and assignments, resulting in lower GPAs and overall rankings. For example, 3.5 GPA students from such talented programs are typically much better than 4.0 GPA students not in those programs.  Similarly, variations in GPA standards exist among different universities, with technical universities generally having lower GPAs than economical universities. These make gaining admission in the US difficult as US faculty are not familiar with these issues.
  \tcblower
  \textbf{Vu}: Vietnamese students and even faculty often lament how this grading system hurts Vietnamese students applying abroad. One way to mitigate this is to make these issues known in your SOP.  \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{Universities with Vietnamese profs} are probably aware of them, but in general your letter writers \emph{and} you can explicitly mention these in their letters and your statement.
\end{commentbox}

@@

\paragraph{How to get TA?}  Unless you have other funding such as RA or fellowships, TA is typically the default for CS PhD programs. In your application, indicate that you need financial assistant. Typically, adcom will either admit you and give you TA, or reject you. We do not admit a student without supporting them.

@@

\paragraph{Ranks} Faculty have ranks such as assistant, associate, and full, regardless if they are tenure-line or teaching.  Assistant means new faculty, associate means they have been promoted, and full means they are senior. Typically, a tenure-track faculty starts with being an assistant professor and then gets tenure and promoted to associate (typically after 6 years) and then full (time varies, some become full within 3--4 years, some 10+ years, some remain associate). A non-tenure track faculty, e.g., teaching or research faculty, also starts with assistant, then is promoted to associated, and full prof.

>>>>

          \item Increasing admission chance by being unique and standing out (\autoref{sec:improve-your-chance}).
          \item It takes about 10 minutes to review an application (\autoref{sec:ievaluate}).
          \item Sometimes adcom members discuss applicants, but in most cases they make independent decisions (\autoref{sec:adcom-discuss}).           
          \item Adcom members only review applications assigned to them (typically matching their expertise) and rarely get involved in other applications (\autoref{sec:read-all}).

@@

In your SOP, focus on research motivation and convince us that you can achieve it through your experience, e.g., published papers. You should also back up your claims with concrete evidence. For example, if you say you have experience with teaching, then show what you did, e.g., undergrad TA or mentoring someone.  If you say you work on a project, then show some results, e.g., paper submitted (or even rejected), achieved X\% performance improvement over ... Also, do not use more than two pages (references can be on the third page). Many SOP samples are \href{https://cs-sop.org/}{available here}.

@@

\begin{commentbox}[Vu:]
  I often read SOP and LORs first. If I am
  persuaded by then, I would skim over other factors and advocate for
  admission (unless I see red flags in other parts). However, if I am not
  convinced, then I will likely recommend rejection (unless I see
  something stand out in other parts).
  \tcblower
  Do careful research on professors, don't mention \emph{emeritus} or adjunct faculty (see more about various types of faculty in \autoref{sec:faculty-types}).
  Also, be careful not to send statements to the wrong schools or mix
  facts (e.g., talking about school X but mentioning working with
  profs. at school Y; and do not talk about George Washington when applying to George Mason). I have seen such statements more times than I should.
\end{commentbox}

@@

\begin{commentbox}[Thanh:]
  Vietnamese universities typically offer specialized programs, such as the talented engineer program at HUST, that have highly competitive entrance exams and a limited number of available slots (e.g., 30 per year). However, these programs often set higher requirements for students, including more demanding tests and assignments, resulting in lower GPAs and overall rankings. For example, 3.5 GPA students from such talented programs are typically much better than 4.0 GPA students not in those programs.  Similarly, variations in GPA standards exist among different universities, with technical universities generally having lower GPAs than economic universities. These make gaining admission in the US difficult as US faculty are not familiar with these issues.
  \tcblower
  \textbf{Vu}: Vietnamese students and even faculty often lament how this grading system hurts Vietnamese students applying abroad. One way to mitigate this is to make these issues known in your SOP.  \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{Universities with Vietnamese profs} are probably aware of them, but in general your letter writers \emph{and} you can explicitly mention these in their letters and your statement.
\end{commentbox}

@@

\paragraph{How to get TA?}  Unless you have other funding such as RA or fellowships, TA is typically the default for CS PhD programs. In your application, indicate that you need financial assistance. Typically, adcom will either admit you and give you TA, or reject you. We do not admit a student without supporting them.

@@

\paragraph{Ranks} Faculty have ranks such as assistant, associate, and full, regardless if they are tenure-line or teaching.  Assistant means new faculty, associate means they have been promoted, and full means they are senior. Typically, a tenure-track faculty starts with being an assistant professor and then gets tenure and promoted to associate (typically after 6 years) and then full (time varies, some become full within 3--4 years, some 10+ years, some remain associate). A non-tenure track faculty, e.g., teaching or research faculty, also starts with Assistant, then is promoted to Associated, and Full prof.
